---
title: "Report_Excersise_1"
author: "Timo Trinidad"
date: "2025-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Importing required libraries: 
```{r}
library(dplyr)
library(lubridate)
library(tidyr)
library(readr)
library(stringr)
library(purrr)
```

### Preprocessing in Excel: 
  - Eliminating string descriptions on the top (deleting first three rows).
  - Harmonising variable names (merging names, which were split over several cells).
  - Processed sheet 1 has been saved seperately as csv.
  
### Loading the data set and cleaning: 

```{r}
setwd("./report_1")
df_soil_carbon_stocks_S1 <- read_delim("C:/Users/timot/OneDrive - Universitaet Bern/Studium/Geographie/2. Aufbaustudium/AGDS I/agds_report_Trinidad_Timo/Report_1/Data/soil_carbon_stocks_S1.csv", 
    delim = "\t", trim_ws = TRUE)

# clean for empty rows#
for( i in 1:nrow(df_soil_carbon_stocks_S1)) {
  if (all(is.na(df_soil_carbon_stocks_S1[i,1:ncol(df_soil_carbon_stocks_S1)]))) {
    df_soil_carbon_stocks_S1 <- df_soil_carbon_stocks_S1[-i,]
}

}
nrow(df_soil_carbon_stocks_S1 )

#filling empty cells in "Experiment" column: 

for (i in 1:(nrow(df_soil_carbon_stocks_S1) - 1)) {
  if (is.na(df_soil_carbon_stocks_S1[i+1,1])) {
    df_soil_carbon_stocks_S1[i+1,1] <- df_soil_carbon_stocks_S1[i,1]
  } else {
    next
  }
}

# Making sure, numbers are stored as numeric objects: 
sapply(df_soil_carbon_stocks_S1, class) # looks okey.

```


### Aggregation by experiment: 

#### Preparation: 
```{r}
# Creating variable for log-response ratio:
df_soil_carbon_stocks_S1$log_response_ration <-  log(df_soil_carbon_stocks_S1$`g C m-2 mean increased CO2`/ df_soil_carbon_stocks_S1$`g C m-2 mean ambient CO2`)

# creating variable indicating phase: 
df_soil_carbon_stocks_S1$Phase <- df_soil_carbon_stocks_S1$`Time (years)`
df_soil_carbon_stocks_S1 <- df_soil_carbon_stocks_S1 |> 
  mutate(Phase = case_when(
    Phase < 3 ~ 1,
    Phase >= 3 & Phase <= 6 ~ 2,
    Phase > 6 ~ 3
  ))

# defining the Phase-variable as factor (and naming it's values):
df_soil_carbon_stocks_S1$Phase <- factor(df_soil_carbon_stocks_S1$Phase, labels = c("early", "mid", "late"))

```

#### Aggregation by experiment and phase through mean: 

```{r}
df_log_response_ratio_aggregate <- df_soil_carbon_stocks_S1 |> group_by(Phase) |> summarise(log_response_ration = mean(log_response_ration))
```

```{r}
knitr::kable(df_log_response_ratio_aggregate)

```

### Interpretation: 

- What is the data about?
Building on the idea of stimulated plant growth through increased atmospheric CO2 concentrations the paper wants to answer if with increased biomass production also the soil C sink increases, which would eventually add to reduce the CO2 in the atmosphere again, building a crucial feedback within the earth system. To answer this question Van Groeningen et al. conduct a meta-analysis over 53 experiments, in which C-levels in soils exposed to increased CO2-levels were compared to soils under "normal" levels (for different depths and time steps).

- What do I expect my analysis to show?
The question regarding whether the new soil C balance under increased CO2 levels is higher or lower than under normal CO2 conditions depends on whether C inputs from plant growth increases more than C loses from increased respiration (therefore increase in C turnover). I assume this to be the case, thus formulating the following working hypothesis: 
Under increased CO2 concentrations C soil concentration increases until a new balance concentration is reached. 

- How do I interpret my results?
Our results show the mean log respone ratios for the three phases aggregated over all experiments. A positive (negative) log-response ratio indicates here a stronger (weaker) increase in soil C under increased CO2 conditions than under normal conditions. For the first 